The Unicode Cookbook \linebreak for Linguists

### 2015-12-10 book/chapters/ipa_background.tex
revise IPA background

Established in 1886, the \textsc{international phonetic association} (henceforth
\textit{Association}) has long maintained a standard alphabet, the
\texsc{international phonetic alphabet} (henceforth \textit{IPA}), which is a
common standard in linguistics to transcribe sounds of spoken languages. It was
first published in 1888 as an international system of phonetic transcription for
oral languages and for pedagogical purposes. It contained phonetic values for
English, French and German. Diacritics for length and nasalization were already
present in this first version, and the same symbols are still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

Originally, the IPA alphabet was a list of symbols with pronunciation examples
from words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels (e.g.\ for manner of consonant
articulation among others `plosives', `nasales', `fricatives', for place of
articulation among others `bronchiales', `laryngales', `labiales' and for vowels
e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge. Thus, similar to notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, this is part of the
Association's mandate. These changes are captured in the recurring revisions to
the IPA.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

@@

From its earliest days the Association aimed to provide ``a separate sign for
each distinctive sound; that is, for each sound which, being used instead of
another, in the same language, can change the meaning of a word''
\citep[27]{IPA1999}. Distinctive sounds became later known as \textsc{phonemes}
and the IPA has developed historically into a notational devise with a strictly
segmental phonemic view. A phoneme is an abstract theoretical notion derived
from an acoustic signal as produced by speakers in the real world. Therefore the
IPA contains a number of theoretical assumptions about speech and how to
transcribe speech in written form.

@@

\noindent Yet, in some situations diacritics can be used to mark contrasts. The
\textit{Handbook} recommends to limit the use of phonemic diacritics to the
following situations:

@@

The interpretation of the IPA symbols in specific usage is not trivial. Although
the articulatory properties of the IPA symbols themselves are rather succinctly
summarised by the normative description in the \textit{Handbook}, it is common
in practical applications that the transcribed symbols do not precisely
represent the phonetic value of the sounds that they represent. So an IPA symbol
<t> in one transcription is not always the same as an IPA <t> in another
transcription (or even within a single transcription). The interpretation of any
particular <t> is mostly a language-specific convention (or even
resource-specific and possibly even context-specific), which -- unfortunately --
is in most cases not made explicit by users of the IPA.

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single ``correct'' transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood.

Basically, the IPA allows for both a `narrow' phonetic transcription and a
`broad' phonological transcription. A `narrow' phonetic transcription may freely
use the all symbols in the IPA chart with direct reference to their phonetic
value, i.e. the transcriber can indicate with the symbols <ŋ͡m> that the
phonetic value of the attested sound is a simultaneous labial and velar closure
which is voiced and contains nasal airflow, independently of the phonemic status
of this sound in the language in question.

@@

In contrast, the basic goal of a phonemic transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεauɔ].
Similarly, the Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

   \item A symbol used represents the phonemic status, and not necessarily the
         precise phonetic realization. So, different transcriptions might use 
         the same symbol for different underlying sounds.
   \item The symbol that is used for a specific phoneme is not necessarily
         unique, so different transcriptions might use different symbols for the
         same underlying sound.
   \item Allophonic variation can be disregarded in phonemic transcription, so
         the same symbol within a single transcription can represent different
         sounds.

@@

In 1989 an IPA revision convention was held in Kiel, Germany. As per other
revisions, there was expansion and revisions to phonetic symbols in the IPA
chart. Notably the marking of tone was extended with the addition of a second
system for marking linguistic tones (the so-called `Chao' tone letters).
Importantly, however, this was also the first revision to address issues of
computational representations for the IPA -- the principles above of which have
had several ramifications that make the interoperability of electronic
linguistic data extremely difficult.

Prior to the Kiel Convention for the modern revision of the IPA in 1989, 
\cite{Wells1987} collected and published practical approaches to coded 
representations of the IPA, which dealt mainly with the assignment of 
characters on the keyboard. The process of assigning standardized ``computer 
codes'' to phonetic symbols was assigned to the Workgroup on Computer 
Coding (henceforth working group) at the Kiel Convention. This working 
group was tasked with \citep{Esling1990,EslingGaylord1993}:

@@

\noindent The identification of IPA symbols with unique identifiers was 
a first step in formalizing the IPA computationally because it would give 
each symbol an unambiguous numerical identifier called an \textsc{IPA Number}. 
The numbering system was to be comprehensive enough to support future revisions 
of the IPA, including symbol specifications and diacritic placement. The 
application of diacritics was also to be made explicit.

Although the Association had never officially approved a set of names 
for the IPA symbols, each IPA symbol received a unique \textsc{IPA Name}. 
Many symbols already had an informal name (or two) used by linguists, but 
consensus on symbol names was growing due to the recent publication of the 
\textit{Phonetic Symbol Guide} \citep{PullumLadusaw1986}. Thus most of the 
IPA symbol names were taken from \cite{PullumLadusaw1986} \citep[31]{IPA1999}.

The working group insightfully decided that the computing-coding convention 
for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
i.e.\ the IPA Number was not meant to be implemented directly in a computer 
encoding. The working group report's declaration includes the explanatory 
remarks \citep[82]{International1989report}:

@@

\noindent Thus the coding convention was not meant to address the engineering 
aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
it was meant to serve as a basis for a communication-interchange standard for 
creating mapping tables from various computer encodings, fonts, phonetic-character-set 
software, etc., to common IPA Numbers, and thus symbols.\footnote{Remember, at 
this time in the late 1980s there was no stable multilingual computing environment. 
But some solution was needed because scholars were increasingly using personal 
computers for their research and many were quickly adopting electronic mail or 
discussion boards like Usenet as a medium for international exchanges. This was 
before the Internet as we know it today. Most of these systems ran on 8-bit 
hardware systems using a 7-bit ASCII character encoding.}

Furthermore, the assignment of computer codes to IPA symbols was meant to 
represent an unbiased formulation. The Association plays the role of an international 
advisory body and it stated that it should not recommend a particular existing 
system of encoding. In fact, during this time there were a number of coding 
systems used, but none of them had a dominant international position. The 
differences between systems were also either too great or too subtle to warrant 
an attempt at combining them \citep{International1989report}.

The working group assigned each IPA symbol to a unique three-digit number, 
i.e.\ an IPA Number. Encoded in this number scheme is information about the 
status of each symbol (see below). The IPA numbers are listed with the IPA 
symbols and they are also illustrated in IPA chart form (see \cite[84]{EslingGaylord1993} 
or \cite[App. 2]{IPA2007}). The numbers were assigned in linear order (e.g.\ 
[p] 101, [b] 102, [t] 103...) following the IPA revision of 1989 and its update in 1996.

The working group made the decision that no IPA symbol, past or present, 
could be ignored. The comprehensive inclusion of all IPA symbols was to 
anticipate the possibility that some symbols might be added, withdrawn, 
or reintroduced into current or future usage. For example, in the 1989 
revision voiceless implosives < ƥ, ƭ, ƈ, ƙ, ʠ > were added; in the 1993 
revision they were removed. Ligatures like < ʧ, ʤ > are included as formerly 
recognized IPA symbols; they are assigned to the ``200 series'' of IPA numbers 
as members of the group of symbols formerly recognized by the IPA. To ensure 
backwards compatibility, legacy IPA symbols would retain an IPA Number and 
an IPA Name for reference purposes. As we discuss below, this decision is 
later reflected in the Unicode Standard; many legacy IPA symbols reside in 
the \textsc{IPA Extensions} block.

The IPA Number is simply expressed as a ``three-digit number numerical 
directory of digit triples'' \cite{}.\footnote{For practical purposes, 
the IPA Number also served as a typesetter's guide to the IPA chart.} 
The numbering scheme specifies three-digit codes, the first digit of which 
indicates the symbol's category \cite{Esling1990,EslingGaylord1993}:

@@

After a symbol is categorized, it is assigned a number sequentially, 
e.g.\ [i] 301, [e] 302, [ɛ] 303. The system allows for the addition 
of new symbols within the various series by appending them, e.g. [ⱱ] 
184. Former or often used but non-IPA symbols for consonants, vowels 
and diacritics are numbered from x99 backwards. For example, the voiceless 
and voiced postalveolar affricates and fricatives < č, ǰ, š, ž > are 
assigned the IPA numbers 299, 298, 297 and 296, respectively, because 
they are not sanctioned IPA symbols.

The assignment of the IPA numbers to IPA symbols provided the basis for 
uniquely identifying the set of past and present IPA symbols as a type of 
computational representational standard of the IPA. Within each revision 
of the IPA, the coding defines a closed and clearly defined set of characters. 
The benefits of this standardization are clear in at least two ways: it is 
used in translation tables that reference ASCII representations of the IPA, 
and this early computational representation of the IPA became the basis for 
its inclusion into the Unicode Standard version 1.0.

\section{SAMPA and X-SAMPA}

@@

For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom made fonts to specific code points.\footnote{For 
example, SIL's popular font SIL IPA 1990}. However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Furthermore, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a commonly encountered example of the non-interoperability of
data and data formats.

One solution was the ASCII-ification of the IPA, which simply involved 
defining keyboard-able sequences as IPA symbol codings.\footnote{\cite{Wells1987} 
provides an in-depth description of IPA codings from country-to-country. 
Later ASCII-IPAs include Kirshenbaum (created in 1992 in a Usenet group and 
named after its lead developer who was at Hewlett-Packard Laboratories) and 
Worldbet (published in 1993 by \cite{Hieronymus1993}, who was at AT\&T Laboratories).} 
A successful effort was SAMPA (Speech Assessment
Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
represent IPA symbols with ASCII
character sequences \citep{Wells1987,Wells_etal1992}, e.g. <p\textbackslash> 
for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
of the principal European Union languages \citep{Wells1995}. It is still widely 
used in language technology.

@@

An extended version of SAMPA, called X-SAMPA, set out to include every symbol, 
including all diacritics, in the IPA chart \citep{Wells1995}. X-SAMPA is considered
more universally applicable because it consists of one table that encodes all
characters in IPA. In line with the principles of the IPA, SAMPA and X-SAMPA include a 
repertoire of symbols. These symbols are intended to represent phonemes rather than 
all allophonic distinctions. Additionally, both ASCII-ifications of IPA are useful because 
strings of SAMPA or X-SAMPA are (reportedly) uniquely parsable \citep{Wells1995}. However, 
like the IPA, X-SAMPA has different notations for encoding the same phonetic phenomena 
(see Pitfall \ref{}).

SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in
computational linguistics. In fact, they are stilled used in popular software packages 
that require ASCII input and some of which have been co-opted for linguistic analyses, 
e.g.~RuG/L04 and SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and 
\url{http://www.splitstree.org/}, respectively}

\section{The need for a single multilingual environment}

During the 1980's, it became increasingly clear that an adequate solution 
to the problem of multilingual computing environments was needed. Linguists 
were on the forefront of addressing this issue because they faced these 
challenges head-on by wishing to publish and communicate electronic text 
with phonetic symbols which were not included in basic ASCII.\footnote{One 
only needs to look at facsimiles of older electronic documents to see exotic 
symbols written in by hand.}

Long familiar were linguists already with the distinction between function 
and form. Even in the context of the computer implementation of writing systems, 
the necessity to distinguish form and function had been made \citep{Becker1984}. 
The computer industry, on the other hand, did not consider, ignored, or simply 
did not encode this principle when creating operating systems like MS-DOS, which 
were limited to 256 code points (due to computer hardware architecture) and 
encoded with one-to-one mappings from character codes to graphemes.

Industry was starting to tackle the issues involved in developing a single 
multilingual computing environment on a variety of fronts, including the then 
new technology of bitmap fonts and the creation of Font Manager and Script 
Manager by Apple \citep{Apple1985,Apple1986,Apple1988}. As noted above, around 
this time linguists were developing work-arounds such as SAMPA, so that they 
could communicate IPA transcription and use ASCII-based software. Some linguists 
formalized the issues of multilingual text processing from a computational 
perspective \citep{Anderson1984,Becker1984,Simons1989}. The study of writing 
systems was also being invigorated by the computational challenges in making 
computers work in a multilingual environment.\footnote{\cite[11--15]{Sampson1985} 
urges linguists to view the study of writing systems as legitimate scientific enquiry.}

@@

The second major benefit of the standardization of the IPA in a computational 
representation by the Kiel working group is that it provided the basis for a 
formal proposal to be submitted to various international standards organizations, 
several of which were trying to tackle (and in a sense `win') the multilingual 
computing environment problem. Basically, everyone from corporations to governments 
to language scientists (for lack of a better term) wanted a single unified multilingual 
character encoding set for all the world's writing systems, even if they did not 
understand or appreciate the challenges involved in creating and adopting a solution. 
Additionally, advancements in computer hardware were making the solutions easier 
to implement in software.

The engineering problems and solutions had been spelled out years before, e.g.\ 
a two-byte encoding for multilingual text \citep{Anderson1984}. 
Although languages vary to an astounding extent (cf.\ \cite{EvansLevinson2009}), 
writing systems are quite similar formally and the issue of formal representation 
of the world's orthographic systems had also been addressed \citep{Simons1989}.

A major obstacle in creating a single encoding multilingual environment from 
the perspective of writing systems involves the distinction between function 
and form \citep{Becker1984} This distinction is so central to basic linguistic 
theory and that trained linguists and semiologists take it as second nature. 
A central challenge in developing a universal character set was to combine a 
technological solution with a formalization of writing systems proper.\footnote{Of 
course there were additional practical issues to overcome, e.g.\ funding, creating 
the formal and technological proposal, deciding which characters and writing systems 
to include initially, while setting precedence of how to add new ones in the future.}

In hindsight it is easy to lose sight of how impactful 30 years of technological 
development have been on linguistics, from theory development using quantitative 
means to pure data collection and dissemination (which as fed back into the former). 
But at the end of the 1970s, virtually no ordinary working linguist was using a 
person computer \cite{Simons1996}. Personal computer usage, however, dramatically 
increased throughout the 1980s. By 1990, dozens of character sets were in common use. 
They varied in their architecture and in their character repertoires, which 
made things a mess. There were two major players in the universal character 
set race: Unicode and the International Organization for Standardization (ISO).

\section{Unicode and ISO 10646}

In the late 1980s, a universal character set was being developed by what 
is now referred to as the Unicode Consortium.\footnote{The Unicode 
Consortium was officially incorporated in January 1991.} This consortium 
consisted largely, although not entirely, of major US corporations, with 
the aim of overcoming the inoperability of different coded character sets 
and their costly hinderance for developing multilingual software development 
and for internationalization efforts. Commercial importance of course drove 
the early inclusion of Latin, non-Latin, and some exotic scripts; see the 
table of commercial importance as measured by GDP of countries using certain 
writing systems \citep[2]{unicode88}.

>>>>

Established in 1886, the \textsc{international phonetic association} (henceforth
\textit{Association}) has long maintained a standard alphabet, the
\textsc{international phonetic alphabet} (henceforth \textit{IPA}), which is a
common standard in linguistics to transcribe sounds of spoken languages. It was
first published in 1888 as an international system of phonetic transcription for
oral languages and for pedagogical purposes. It contained phonetic values for
English, French and German. Diacritics for length and nasalization were already
present in this first version, and the same symbols are still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

Originally, the IPA was a list of symbols with pronunciation examples
using words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels (e.g.\ for manner of
articulation among others `plosives', `nasales', `fricatives', for place of
articulation among others `bronchiales', `laryngales', `labiales' and for vowels
e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge. Thus, similar to notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, this is part of the
Association's mandate. These changes are captured in the recurring revisions to
the IPA.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

@@

\noindent From its earliest days the Association aimed to provide ``a separate sign for
each distinctive sound; that is, for each sound which, being used instead of
another, in the same language, can change the meaning of a word''
\citep[27]{IPA1999}. Distinctive sounds became later known as \textsc{phonemes}
and the IPA has developed historically into a notational devise with a strictly
segmental phonemic view. A phoneme is an abstract theoretical notion derived
from an acoustic signal as produced by speakers in the real world. Therefore the
IPA contains a number of theoretical assumptions about speech and how to
transcribe speech in written form.

@@

\noindent Yet, in some situations diacritics can be used to mark phonemic
contrasts. The \textit{Handbook} recommends to limit the use of phonemic
diacritics to the following situations:

@@

The interpretation of the IPA symbols in specific usage is not trivial. Although
the articulatory properties of the IPA symbols themselves are rather succinctly
summarised by the normative description in the \textit{Handbook}, it is common
in practical applications that the transcribed symbols do not precisely
represent the phonetic value of the sounds that they represent. So an IPA symbol
<t> in one transcription is not always the same as an IPA <t> in another
transcription (or even within a single transcription). The interpretation of any
particular <t> is mostly a language-specific convention (or even
resource-specific and possibly even context-specific), a fact which --
unfortunately -- is in most cases not made explicit by users of the IPA.

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single ``correct'' transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood. Basically, the IPA allows for both a `narrow' phonetic transcription and a
`broad' phonological transcription. A `narrow' phonetic transcription may freely
use the all symbols in the IPA chart with direct reference to their phonetic
value, i.e. the transcriber can indicate with the symbols <ŋ͡m> that the
phonetic value of the attested sound is a simultaneous labial and velar closure
which is voiced and contains nasal airflow, independently of the phonemic status
of this sound in the language in question.

@@

In contrast, the basic goal of a phonemic ``broad'' transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεauɔ].
Similarly, the Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

   \item A symbol represents the phonemic status, and not necessarily the
         precise phonetic realization. So, different transcriptions might use 
         the same symbol for different underlying sounds.
   \item Any symbol that is used for a specific phoneme is not necessarily
         unique, so different transcriptions might use different symbols for the
         same underlying sound.
   \item Allophonic variation can be disregarded in phonemic transcription, so
         the same symbol within a single transcription can represent different
         sounds.

@@

In 1989 an IPA revision convention was held in Kiel, Germany. As per other
revisions, there was expansion and revisions to phonetic symbols in the IPA
chart. Notably the marking of tone was extended with the addition of a second
system for marking linguistic tones (the so-called `Chao' tone letters).
Importantly, however, this was also the first revision to address issues of
computational representations for the IPA.

Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized ``computer
codes'' to phonetic symbols was assigned to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

\noindent The identification of IPA symbols with unique identifiers was 
a first step in formalizing the IPA computationally because it would give 
each symbol an unambiguous numerical identifier called an \textsc{IPA number}. 
The numbering system was to be comprehensive enough to support future revisions 
of the IPA, including symbol specifications and diacritic placement. The 
application of diacritics was also to be made explicit.

Although the Association had never officially approved a set of names 
for the IPA symbols, each IPA symbol received a unique \textsc{IPA name}. 
Many symbols already had an informal name (or two) used by linguists, but 
consensus on symbol names was growing due to the recent publication of the 
\textit{Phonetic Symbol Guide} \citep{PullumLadusaw1986}. Thus most of the 
IPA symbol names were taken from that source \citep[31]{IPA1999}.

The working group insightfully decided that the computing-coding convention 
for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
i.e.\ the IPA Number was not meant to be implemented directly in a computer 
encoding. The working group report's declaration includes the following explanatory 
remarks \citep[82]{International1989report}:

@@

\noindent Thus the coding convention was not meant to address the engineering 
aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
it was meant to serve as a basis for a communication-interchange standard for 
creating mapping tables from various computer encodings, fonts, phonetic-character-set 
software, etc., to common IPA Numbers, and thus symbols.\footnote{Remember, at 
this time in the late 1980s there was no stable multilingual computing environment. 
But some solution was needed because scholars were increasingly using personal 
computers for their research and many were quickly adopting electronic mail or 
discussion boards like Usenet as a medium for international exchanges. Most of these systems ran on 8-bit 
hardware systems using a 7-bit ASCII character encoding.}

Furthermore, the assignment of computer codes to IPA symbols was meant to
represent an unbiased formulation. The Association here played the role of an
international advisory body and it stated that it should not recommend a
particular existing system of encoding. In fact, during this time there were a
number of coding systems used (see Section~\ref{encoding}), but none of them had
a dominant international position. The differences between systems were also
either too great or too subtle to warrant an attempt at combining them
\citep{International1989report}.

The working group assigned each IPA symbol to a unique three-digit number, i.e.\
an IPA Number. Encoded in this number scheme implicitly includes information
about the status of each symbol (see below). The IPA numbers were listed with the
IPA symbols and they were also illustrated in IPA chart form (see
\cite[84]{EslingGaylord1993} or \cite[App. 2]{IPA2007}). The numbers were
assigned in linear order (e.g.\ [p] 101, [b] 102, [t] 103...) following the IPA
revision of 1989 and its update in 1996. Although the numbering scheme still 
exists, in practice is it superseded by the Unicode codification of symbols.

The working group made the decision that no IPA symbol, past or present, 
could be ignored. The comprehensive inclusion of all IPA symbols was to 
anticipate the possibility that some symbols might be added, withdrawn, 
or reintroduced into current or future usage. For example, in the 1989 
revision voiceless implosives <~ƥ,~ƭ,~ƈ,~ƙ,~ʠ~> were added; in the 1993 
revision they were removed. Ligatures like <~ʧ,~ʤ~> are included as formerly 
recognized IPA symbols; they are assigned to the ``200 series'' of IPA numbers 
as members of the group of symbols formerly recognized by the IPA. To ensure 
backwards compatibility, legacy IPA symbols would retain an IPA Number and 
an IPA Name for reference purposes. As we discuss below, this decision is 
later reflected in the Unicode Standard as many legacy IPA symbols still reside in 
the \textsc{IPA extensions} block.

The IPA Number is simply expressed as a ``three-digit number numerical directory
of digit triples'' \cite{}. The numbering scheme specifies three-digit codes,
the first digit of which indicates the symbol's category
\citep{Esling1990,EslingGaylord1993}:

@@

After a symbol is categorized, it is assigned a number sequentially, e.g.\ [i]
301, [e] 302, [ɛ] 303. The system allows for the addition of new symbols within
the various series by appending them, e.g. [\charis{ⱱ}] 184. Former non-IPA
symbols (or often-used but non-offical IPA symbols) for consonants, vowels and
diacritics are numbered from x99 backwards. For example, the voiceless and
voiced postalveolar affricates and fricatives <~č,~ǰ,~š,~ž~> are assigned the
IPA numbers 299, 298, 297 and 296, respectively, because they are not sanctioned
IPA symbols.

The assignment of the IPA numbers to IPA symbols provided the basis for uniquely
identifying the set of past and present IPA symbols as a type of computational
representational standard of the IPA. Within each revision of the IPA, the
coding defines a closed and clearly defined set of characters. The benefits of
this standardization are clear in at least two ways: it is used in translation
tables that reference ASCII representations of the IPA, and this early
computational representation of the IPA became the basis X-SAMPA and for the
inclusion of the IPA into the Unicode Standard version 1.0.

% ==========================
\section{SAMPA and X-SAMPA}
% ==========================

@@

For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A more recent survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom made fonts to specific code points.\footnote{For 
example, SIL's popular font SIL IPA 1990}. However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Furthermore, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a commonly encountered example of the non-interoperability of
data and data formats.

One solution was the ASCII-ification of the IPA, which simply involved 
defining keyboard-able sequences consisting of ASCII combinations as an IPA symbol
codings. For example, \cite{Wells1987} provides an in-depth description of IPA
codings from country-to-country. Later ASCII-IPAs include Kirshenbaum (created
in 1992 in a Usenet group and named after its lead developer who was at
Hewlett-Packard Laboratories) and Worldbet (published in 1993 by
\cite{Hieronymus1993}, who was at AT\&T Laboratories). 
The most successful effort was SAMPA (Speech Assessment
Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
represent IPA symbols with ASCII
character sequences \citep{Wells1987,Wells_etal1992}, using e.g. <p\textbackslash> 
for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
of the principal European languages \citep{Wells1995}. It is still widely 
used in language technology.

@@

An extended version of SAMPA, called X-SAMPA, set out to include every symbol,
including all diacritics, in the IPA chart \citep{Wells1995}. X-SAMPA is
considered more universally applicable because it consists of one table that
encodes all characters in IPA. In line with the principles of the IPA, SAMPA and
X-SAMPA include a repertoire of symbols. These symbols are intended to represent
phonemes rather than all allophonic distinctions. Additionally, both
ASCII-ifications of IPA -- SAMPA and X-SAMPA -- are
(reportedly) uniquely parsable \citep{Wells1995}. However, like the IPA, X-SAMPA
has different notations for encoding the same phonetic phenomena (cf. Section~
\ref{pitfall-multiple-options-ipa}).

SAMPA and X-SAMPA have been widely used for speech technology and as an encoding
system in computational linguistics. In fact, they are still used in popular
software packages that require ASCII input, e.g.~RuG/L04 and SplitsTree4.\footnote{See
\url{http://www.let.rug.nl/kleiweg/L04/} and \url{http://www.splitstree.org/},
respectively}

% ==========================
\section{The need for a single multilingual environment}
% ==========================

In hindsight it is easy to lose sight of how impactful 30 years of technological
development have been on linguistics, from theory development using quantitative
means to pure data collection and dissemination. But at the end of the 1970s,
virtually no ordinary working linguist was using a personal computer
\citep{Simons1996}. Personal computer usage, however, dramatically increased
throughout the 1980s. By 1990, dozens of character sets were in common use. They
varied in their architecture and in their character repertoires, which made
things a mess.

During the 1980's, it became increasingly clear that an adequate solution 
to the problem of multilingual computing environments was needed. Linguists 
were on the forefront of addressing this issue because they faced these 
challenges head-on by wishing to publish and communicate electronic text 
with phonetic symbols which were not included in basic ASCII. One 
only needs to look at facsimiles of older electronic documents to see exotic 
symbols written in by hand after the preparation of typed version.

%There were two major players in the universal character set race:
%Unicode and the International Organization for Standardization (ISO).

%Long familiar were linguists already with the distinction between function 
%and form. Even in the context of the computer implementation of writing systems, 
%the necessity to distinguish form and function had been made \citep{Becker1984}. 
%The computer industry, on the other hand, did not consider, ignored, or simply 
%did not encode this principle when creating operating systems like MS-DOS, which 
%were limited to 256 code points (due to computer hardware architecture) and 
%encoded with one-to-one mappings from character codes to graphemes.

@@

A major benefit of the standardization of the IPA in a computational
representation by the Kiel working group is that it provided the basis for a
formal proposal to be submitted to various international standards
organizations, several of which were trying to tackle (and in a sense `win') the
multilingual computing environment problem (cf. Section~\ref{encoding}).
Basically, everyone -- from corporations to governments to language scientists
-- wanted a single unified multilingual character encoding set for all the
world's writing systems, even if they did not understand or appreciate the
challenges involved in creating and adopting a solution.

Industry was starting to tackle the issues involved in developing a single
multilingual computing environment on a variety of fronts, including the then
new technology of bitmap fonts and the creation of Font Manager and Script
Manager by Apple \citep{Apple1985,Apple1986,Apple1988}. As noted above, around
this time linguists were developing work-arounds such as SAMPA, so that they
could communicate IPA transcription and use ASCII-based software. Some linguists
formalized the issues of multilingual text processing from a computational
perspective \citep{Anderson1984,Becker1984,Simons1989}. The study of writing
systems was also being invigorated \citep[11--15]{Sampson1985} by the
computational challenges in making computers work in a multilingual environment.
The engineering problems and solutions had been spelled out years before, e.g.\
a two-byte encoding for multilingual text \citep{Anderson1984}. Although
languages vary to an astounding extent (cf.\ \cite{EvansLevinson2009}), writing
systems are quite similar formally and the issue of formal representation of the
world's orthographic systems had already been addressed \citep{Simons1989}.

%A major obstacle in creating a single encoding multilingual environment from 
%the perspective of writing systems involves the distinction between function 
%and form \citep{Becker1984} This distinction is so central to basic linguistic 
%theory and that trained linguists and semiologists take it as second nature. 
%A central challenge in developing a universal character set was to combine a 
%technological solution with a formalization of writing systems proper.\footnote{Of 
%course there were additional practical issues to overcome, e.g.\ funding, creating 
%the formal and technological proposal, deciding which characters and writing systems 
%to include initially, while setting precedence of how to add new ones in the future.}

After the Kiel Convention in 1989, the working group assisted
the International Phonetic Association in representing the IPA to the
\textsc{international standards organisation} (ISO) and to the \textsc{text
encoding initiative} (TEI) \citep{EslingGaylord1993}. The working group's
formalization of the IPA, i.e.\ a full listing of agreed upon computer codings
for phonetic symbols, was used in developing writing system descriptions, which
were at the time being solicited for scripts to be included in the new
multilingual international character encoding standards. The working group for
ISO/IEC 10646 and Unicode were two such initiatives.

In the historical context of the IPA being considered for inclusion in 
ISO/IEC 10646, it is important to realize that there were a variety of 
sources (i.e.\ not just from the Association) which submitted character 
proposals for phonetic alphabets. These proposals, including the one from the 
Association via the Kiel working group, were considered as a whole by 
the ISO working groups that were responsible for incorporating a phonetic 
script into the universal character set (UCS). The ISO working groups that 
were responsible for assigning a phonetic character set then made their 
own submissions as part of a review process by ISO for approval based on 
both ``informatic'' and phonetic criteria \citep[86]{EslingGaylord1993}.

Character set ISO/IEC 10646 was approved by ISO, including the phonetic
characters submitted to them in May 1993. The set of IPA characters were
assigned UCS codes in 16 bit representation (in hexadecimal) and were published
Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
representation of the IPA Symbol, its IPA Name, phonetic description, IPA
Number, UCS Code and AFII Code.\footnote{The \textsc{association for font
information interchange} (AFII) was an international database of glyphs created
to promote the standardization of font data required to produce ISO/IEC 10646.} When the
character sets of ISO/IEC 10646 and the Unicode Standard later converged (see
Section~\ref{the-unicode-approach}), the IPA proposal was
included in the Unicode Standard Version 1.0 -- largely as we know it
today.
